Meta-analysis of gut microbiome studies identifies disease-specific and shared responses

### 2018-12-10 chap1.tex
Small changes

The microbes that live in and on our bodies make up the human microbiome, are essential for health, and have been implicated in many diseases.
Almost all human body sites are colonized by microbes, ranging from the gut, the largest human-associated microbial community, to the lungs, which were for many years considered sterile \cite{sender-2016-bhratio,beck-2012-lungmicrobiome}.
These microbial communities perform essential functions for health, including fighting off and preventing infections, regulating host metabolism and interacting with the immune system, and metabolizing xenobiotics or other compounds which are indigestible by the host.
Additionally, perturbations in human microbiomes have been implicated in many diseases, including inflammatory, metabolic, neurological, and respiratory conditions \cite{beck-2012-lungmicrobiome,ibd-papa,ob-TBD,hsiao-2013-autism}.

>>>>

The microbes that live in and on our bodies make up the human microbiome, are essential for health, and have been implicated in many diseases.
Almost all human body sites are colonized by microbes, ranging from the gut, the largest human-associated microbial community, to the lungs, which were for many years considered sterile \cite{sender-2016-bhratio,beck-2012-lungmicrobiome}.
These microbial communities perform essential functions for health, including fighting off and preventing infections, regulating host metabolism and interacting with the immune system, and metabolizing xenobiotics or other compounds which are indigestible by the host.
Additionally, perturbations in human microbiomes have been implicated in many diseases, including inflammatory, metabolic, neurological, and respiratory conditions \cite{beck-2012-lungmicrobiome,ibd-papa,ridaura-2013-obfmt,hsiao-2013-autism}.

### 2018-12-10 chap5.tex
Small changes

A core theme that has emerged from my thesis work is that re-analyzing existing microbiome datasets can add substantial value to our field.
One of my personal conclusions from the meta-analysis is that cross-sectional gut microbiome studies should be regularly contextualized within the existing published body of work.
Now that microbiome datasets are more readily available and bioinformatics tools to process and analyze them are becoming very accessible, I hope that researchers make it a habit to ask: are these associations consistent with ones from independent patient cohorts? Are they specific to my disease of interest or are they part of a non-specific response to health and disease?

I also found the value-add of re-analyzing datasets especially relevant in microbiome research led by clinicians.
For example, the original IBD FMT trials that I re-analyzed in Chapter 4 did not thoroughly investigate potential donor effects on patient response in their original studies.
This is expected in part because heterogenous donor response was not usually their primary question of interest, and also because their trials were not powered for these analyses.
However, now that multiple IBD FMT trials have been published with patient and donor microbiome sequencing data, hypotheses about what leads to improved patient response could be tested \textit{in silico}, as we did for butyrate producer abundance in Chapter 4.
In Chapter 3, many of my datasets were pulled from studies led by clinical researchers who may not have had access to the bioinformatics and statistical expertise to fully interrogate disease associations within their datasets.
By making their data publicly available, their work continued to contribute new knowledge to the field.
% maybe: Finally, re-analyzing published datasets adds value at little extra cost: recruiting patients, collecting samples, and generating data

Moving forward, I hope and expect that researchers continue to make their raw data publicly available and that re-analyses of such data become standard in the field.
Publicly available raw data allows researchers to ask and answer new questions, testing their hypotheses \textit{in silico} without the need for costly new studies.
New studies can also compare their results with existing work to determine which of their findings hold across different studies and which are specific to their specific study.
However, using published data in new contexts comes with challenges.
Batch effects, which result from different experimental and sequencing methods, can make it very difficult to compare data across different labs.
As a follow-up to the meta-analysis, I contributed to a method to correct for batch effects in case-control studies \cite{gibbons-2018}, and I hope that batch-correction methods keep being developed for this data type.
Another major challenge is data availability, and specifically clinical metadata like patient disease diagnosis.
Through this work, I have come to realize that raw data without its associated metadata is in almost all cases useless.
Given these challenges, standards for sharing data which respect patient privacy, clinicians' efforts for patient recruitment, and the needs of computational biologists will need to be agreed upon and upheld as a community.

>>>>

A core theme that has emerged from my thesis work is that re-analyzing existing microbiome datasets can add substantial value to our field.
One of my personal conclusions from the meta-analysis is that new cross-sectional gut microbiome studies should be regularly contextualized within the existing published body of work, almost as a mini meta-analysis of their own.
Now that microbiome datasets are more readily available and bioinformatics tools to process and analyze them are becoming very accessible, I hope that researchers make it a habit to ask: are these associations replicated in independent patient cohorts? And: are they specific to my disease of interest or are they part of a non-specific response to health and disease?

I also found the value-add of re-analyzing datasets especially relevant in microbiome research led by clinicians.
For example, the original IBD FMT trials that I re-analyzed in Chapter 4 did not thoroughly investigate potential donor effects on patient response in their original studies.
This is expected in part because donor heterogeneity was not usually a primary research question of interest, and also because the trials were not powered for such analyses.
However, now that multiple IBD FMT trials have been published with paired patient and donor microbiome sequencing data, hypotheses about what leads to improved patient response can be tested \textit{in silico} (as we did for butyrate producer abundance in Chapter 4).
In Chapter 3, many of my datasets were pulled from studies led by clinical researchers who may not have had access to the bioinformatics and statistical expertise to fully interrogate disease associations within their datasets.
By making their data publicly available, their work continued to contribute new knowledge to the field.
% maybe: Finally, re-analyzing published datasets adds value at little extra cost: recruiting patients, collecting samples, and generating data

Moving forward, I hope and expect that researchers continue to make their raw data publicly available and that re-analyses of such data become standard in the field.
Publicly available raw data allows researchers to ask and answer new questions, testing their hypotheses \textit{in silico} without the need for costly new studies.
New studies can also be compared with existing work to determine which of their findings hold across different studies and which are specific to their specific study.
However, using published data in new contexts comes with challenges.
Batch effects, which result from different experimental and sequencing methods, can make it very difficult to compare data across different labs.
As a follow-up to the meta-analysis, I contributed to a method to correct for batch effects in case-control studies \cite{gibbons-2018}, and I hope that batch-correction methods keep being developed for this data type.
Another major challenge is data availability, and specifically clinical metadata like patient disease diagnosis.
Through this work, I have come to realize that raw data without its associated metadata is in almost all cases useless.
Given these challenges, standards for sharing data which respect patient privacy, clinicians' efforts for patient recruitment, and the needs of computational biologists will need to be agreed upon and upheld as a community.

### 2018-12-10 cover.tex
Start integrating aspiration paper:
Small changes

\author{Claire Marie Noëlle Duvallet}

@@

Thomas - being my first introduction to computational work with patience and enthusiasm, enabling me to find my passion here
Mariana - drive and ability to make impossible things actually happen was an inspiration, expanded the scope of what a PhD student could do
Scott - being a role model, mentor, collaborator, and now close friend. Hope you remain all four forever. Also for encouraging me to start doing diversity and REFS
Sean Gibbons - modeling pure interest in science for the sake of science, encouraging me and being patient when I thought it was all fake, modeling success and being a joy to work with

>>>>

\author{Claire Marie Noelle Duvallet}

@@

Thomas - being my first introduction to computational work with patience and enthusiasm, enabling me to find my passion here
Mariana - drive and ability to make impossible things actually happen was an inspiration, expanded the scope of what a PhD student could do
Scott - being a role model, mentor, collaborator, and now close friend. Hope you remain all four forever. Also for encouraging me to start doing diversity and REFS
Sean Gibbons - modeling pure interest in science for the sake of science, encouraging me and being patient when I thought it was all fake, modeling success and being a joy to work with, constant sponsor of my work (opened a really important door into the world of open-source software)

### 2018-12-10 meta-analysis/manuscript.tex
Small changes

To answer questions about the reproducibility and generalizability of reported associations between the human microbiome and disease, we collected, re-processed, and re-analyzed raw data from a collection of microbiome datasets.
We included studies with publicly available 16S amplicon sequencing data (i.e. FASTQ or FASTA) for stool samples from at least 15 case patients which also had associated disease metadata (i.e. case or control disease labels).
Studies which exclusively focused on children under 5 years old were excluded from our analyses.
We identified over 50 suitable case-control 16S datasets, of which 28 were successfully downloaded, processed, and included in a publicly available database which we called MicrobiomeHD\cite{microbiomehd}.
Characteristics of these datasets, including sample sizes, diseases and conditions, and references, are shown in Table \ref{tab:datasets} and Supplementary Table \ref{tab:datasets_full_info}.
For each downloaded study, we processed the raw sequencing data through our 16S processing pipeline (\url{https://github.com/thomasgurry/amplicon_sequencing_pipeline}) (see Supplementary Tables \ref{tab:processing} and \ref{tab:data} for detailed data sources and processing methods).
100\% denovo OTUs were assigned taxonomy with the RDP classifier \cite{wang2007naive} (c $=$ 0.5), converted to relative abundances by dividing by total sample reads, and collapsed to the genus level. OTUs which were not assigned at the genus level were discarded.
By collapsing data to the genus level, we lost the sensitivity to detect fine-scale differences in species or strain abundances across case and control groups, but we minimized certain batch effects that plague comparisons across studies.
Thus, we took a course-grained approach to optimize our ability to compare data across studies at the expense of phylogenetic resolution.

@@

Some conditions are characterized by a broad restructuring of gut microbial communities.
In these cases, full community restoration strategies like fecal microbiota transplants may be more appropriate.
For example, diarrhea consistently results in large-scale rearrangements in the composition of the gut microbiome, which is likely reflective of reduced stool transit time (Figures \ref{fig:fig1} and \ref{fig:fig2}).
We saw many microbes consistently associated with both \textit{Clostridium difficile} infection (CDI) and non-CDI diarrhea (Figures \ref{fig:fig2} and \ref{fig:fig3}A) \cite{cdi-youngster,cdi-schubert,cdi-vincent,edd-singh}.
In general, Proteobacteria increase in prevalence in patients with diarrhea, with a concomitant decrease in the relative abundances of Bacteroidetes and some Firmicutes.
In particular, we see a reduction in butyrate-producing Clostridia, including genera within \textit{Ruminococcaceae} and \textit{Lachnospiraceae} families, which have been associated with a healthy gut \cite{wong2006colonic}.
We also see an increase in prevalence of genera that contain organisms often associated with lower pH and higher oxygen levels of the upper-gut, like \textit{Lactobacillaceae} and \textit{Enterobacteriaceae}, in patients with diarrhea (Figure \ref{fig:fig3}A) \cite{donaldson2016gut}.
Additionally, both CDI and non-CDI diarrhea patients had lower alpha diversity, a measure of overall community structure, than healthy controls in all studies (Supplementary Figures \ref{fig:fig_alpha}-\ref{fig:fig_alpha_simpson}).
Consistent with the CDI and non-CDI diarrheal studies, we also found that organisms associated with the upper gut, like \textit{Lactobacillus} and \textit{Enterobacteriaceae}, appear to be enriched in IBD patients, who can present with diarrheal symptoms (Supplementary Figure \ref{fig:supp_dis_specific}) \cite{donaldson2016gut,Kirsner1982ibd}.
IBD patients also tended to have lower alpha diversities than controls (Crohn's disease vs. controls in three studies, ulcerative colitis vs. controls in two studies; Supplementary Figures \ref{fig:fig_alpha}-\ref{fig:fig_alpha_simpson}), though this difference was less drastic than in the diarrheal studies where all patients had active diarrhea.

@@

Finally, we sought to determine whether a unified microbiome response to general health and disease could be identified.
Previous studies have proposed that reduced alpha diversity is a reliable indicator of disease-associated dysbiosis \cite{ob-gordon,cdi-vincent,mosca2016gut}.
In our re-analysis, we found no consistent reduction of alpha diversity in case patients, with the exception of diarrhea and perhaps IBD (Supplementary Figures \ref{fig:fig_alpha}-\ref{fig:fig_alpha_simpson}).
These results are consistent with previous meta-analyses, which found inconsistent relationships between alpha diversity and disease and very small effect sizes in non-diarrheal diseases \cite{walters2014meta, Sze07092016}.
To further address the question of whether we could find a robust, generalized signal for diseased microbiomes regardless of the disease type, we built random forest classifiers to distinguish healthy patients from any type of case patient.
The AUCs from these general healthy vs. disease classifiers correlated strongly with the original single-dataset classification results, indicating that there is indeed a general microbiome signal that can be identified even across different diseases (see Supplementary Note \ref{sec:overall_classifier} and Supplementary Figure \ref{fig:overall_classifier}).

@@

Meta-analysis identified potential false positives and false negatives across studies and conditions.
For example, we found that reported associations between alpha diversity and disease within individual studies tended to lose significance when looking across studies, except in the case of diarrhea and perhaps IBD (Supplementary Figures \ref{fig:fig_alpha}-\ref{fig:fig_alpha_simpson}).
Another example of a potential false positive was the association between \textit{Prevotella} and disease.
Autism \cite{asd-kb}, rheumatoid arthritis \cite{ra-littman}, and HIV \cite{lozupone2013alterations,hiv-dinh} have each been reported to be associated with \textit{Prevotella}. For each of these diseases, the associations with \textit{Prevotella} were weakly significant or complicated by confounding factors.
In our more statistically conservative re-analysis, we found no association between autism or arthritis and \textit{Prevotella}.
As mentioned previously, in the case of HIV, the association with \textit{Prevotella} was due to demographic factors unrelated to disease \cite{noguera2016gut}.
Regardless of whether shifts in \textit{Prevotella} are truly biologically related to each studied disease state, it is clear that such shifts are not specific to one particular condition and should not be reported as putative disease-specific biomarkers.
We also found that certain signals picked out by meta-analysis did not always hold within individual studies.
For example, studies with small sample sizes often had few or no significant associations (e.g. Vincent et al. (2014) \cite{cdi-vincent}, Chen et al. (2012) \cite{crc-xiang}, and Willing et al. (2009) \cite{ibd-engstrand}).
Here, the fact that other studies analyzing the same diseases consistently found associations strengthens the hypothesis that the lack of microbiome-associated signal in these studies was due to low power rather than a lack of true signal.
Because individual studies are plagued by low statistical power, confounding variables, and batch effects which can obscure biological signals, the identification of disease-specific and non-specific microbial associations will continue to improve as more datasets and diseases are included in future meta-analyses.

@@

\subsection*{16S processing}
Raw data were downloaded and processed through our in-house 16S processing pipeline ({\url{ https://github.com/thomasgurry/amplicon_sequencing_pipeline}).
Data and metadata were acquired as described in Supplementary Table \ref{tab:data}.
When needed, we de-multiplexed sequences by finding exact matches to the provided barcodes and trimmed primers with a maximum of 1 mismatch.
In general, sequences were quality filtered by truncating at the first base with Q $<$ 25.
However, some datasets did not pass this stringent quality threshold (i.e. the resulting OTU table was either missing many of the original samples, or the read depth was significantly lower than reported in the original paper).
For 454 data, we loosened the quality threshold to 20, whereas for paired-end Illumina data we removed reads with more than 2 expected errors.
If possible, all reads were trimmed to 200 bp.
In cases where this length trimming discarded a majority of sequences, we lowered our threshold to 150 or 101 bp.
The specific processing parameters we used for each dataset can be found in Supplementary Table \ref{tab:processing}.
To assign OTUs, we clustered OTUs at 100\% similarity using USEARCH \cite{edgar-usearch-2010} and assigned taxonomy to the resulting OTUs with the RDP classifier \cite{wang2007naive} and a confidence cutoff of 0.5.
For each dataset, we removed samples with fewer than 100 reads and OTUs with fewer than 10 reads, as well as OTUs which were present in fewer than 1\% of samples within a study.
We calculated the relative abundance of each OTU by dividing its value by the total reads per sample.
We then collapsed OTUs to genus level by summing their respective relative abundances, discarding any OTUs which were unannotated at the genus level.
All statistical analyses were performed on this genus-level relative abundance data.

@@

\section{Figures}

\newpage
\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=1\textwidth]{samplesize_auc_nsig_balance.pdf}}
    \captionsetup{font=footnotesize,labelfont=footnotesize}
	\caption{\textbf{Most diseases show microbiome alterations, and consistent disease-associated shifts differ in their extent and direction.} (A) Left: Total sample size for each study included in these analyses. Additional information about each dataset can be found in Table \ref{tab:datasets}. Studies on the y-axis are grouped by disease and ordered by decreasing sample size (top to bottom). Right: Area under the ROC curve (AUC) for genus-level random forest classifiers. X-axis starts at 0.5, the expected value for a classifier which assigns labels randomly, and AUCs less than 0.5 are not shown. ROC curves for all datasets are in Supplementary Figure \ref{fig:roc_curves}. Note that Youngster et al. (2014) \cite{cdi-youngster} had only 4 distinct control patients was excluded from the Random Forest analysis. (B) Left: Number of genera with q $<$ 0.05 (Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction) for each dataset. If a study has no significant associations, no point is shown. Right: Direction of the microbiome shift, i.e. the percent of total associated genera which were enriched in diseased patients. In datasets on the leftmost blue line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are health-associated (i.e. depleted in patients relative to controls). In datasets on the rightmost red line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are disease-associated (i.e. enriched in patients relative to controls). Supplementary Figures \ref{fig:overall_heatmap_qvalues} and \ref{fig:overall_heatmap_foldchange} show q-values and effects for each genus in each study.}
	\label{fig:fig1}
	\end{center}
\end{figure}

>>>>

To answer questions about the reproducibility and generalizability of reported associations between the human microbiome and disease, we collected, re-processed, and re-analyzed raw data from a collection of microbiome datasets.
We included studies with publicly available 16S amplicon sequencing data (i.e. FASTQ or FASTA) for stool samples from at least 15 case patients which also had associated disease metadata (i.e. case or control disease labels).
Studies which exclusively focused on children under 5 years old were excluded from our analyses.
We identified over 50 suitable case-control 16S datasets, of which 28 were successfully downloaded, processed, and included in a publicly available database which we called MicrobiomeHD \cite{microbiomehd}.
Characteristics of these datasets, including sample sizes, diseases and conditions, and references, are shown in Table \ref{tab:datasets} and Supplementary Table \ref{tab:datasets_full_info}.
For each downloaded study, we processed the raw sequencing data through our 16S processing pipeline (\url{https://github.com/thomasgurry/amplicon_sequencing_pipeline}) (see Supplementary Tables \ref{tab:processing} and \ref{tab:data} for detailed data sources and processing methods).
100\% denovo OTUs were assigned taxonomy with the RDP classifier \cite{wang2007naive} (c $=$ 0.5), converted to relative abundances by dividing by total sample reads, and collapsed to the genus level. OTUs which were not assigned at the genus level were discarded.
By collapsing data to the genus level, we lost the sensitivity to detect fine-scale differences in species or strain abundances across case and control groups, but we minimized certain batch effects that plague comparisons across studies.
Thus, we took a course-grained approach to optimize our ability to compare data across studies at the expense of phylogenetic resolution.

@@

Some conditions are characterized by a broad restructuring of gut microbial communities.
In these cases, full community restoration strategies like fecal microbiota transplants may be more appropriate.
For example, diarrhea consistently results in large-scale rearrangements in the composition of the gut microbiome, which is likely reflective of reduced stool transit time (Figures \ref{fig:fig1} and \ref{fig:fig2}).
We saw many microbes consistently associated with both \textit{Clostridium difficile} infection (CDI) and non-CDI diarrhea (Figures \ref{fig:fig2} and \ref{fig:fig3}A) \cite{cdi-youngster,cdi-schubert,cdi-vincent,edd-singh}.
In general, Proteobacteria increase in prevalence in patients with diarrhea, with a concomitant decrease in the relative abundances of Bacteroidetes and some Firmicutes.
In particular, we see a reduction in butyrate-producing Clostridia, including genera within \textit{Ruminococcaceae} and \textit{Lachnospiraceae} families, which have been associated with a healthy gut \cite{wong2006colonic}.
We also see an increase in prevalence of genera that contain organisms often associated with lower pH and higher oxygen levels of the upper-gut, like \textit{Lactobacillaceae} and \textit{Enterobacteriaceae}, in patients with diarrhea (Figure \ref{fig:fig3}A) \cite{donaldson2016gut}.
Additionally, both CDI and non-CDI diarrhea patients had lower alpha diversity, a measure of overall community structure, than healthy controls in all studies (Supplementary Figures \ref{fig:fig_alpha}-\ref{fig:fig_alpha_simpson}).
Consistent with the CDI and non-CDI diarrheal studies, we also found that organisms associated with the upper gut, like \textit{Lactobacillus} and \textit{Enterobacteriaceae}, appear to be enriched in IBD patients, who can present with diarrheal symptoms (Supplementary Figure \ref{fig:supp_dis_specific}) \cite{donaldson2016gut,Kirsner1982ibd}.
IBD patients also tended to have lower alpha diversities than controls (Crohn's disease vs. controls in three studies, ulcerative colitis vs. controls in two studies; Supplementary Figures \ref{fig:fig_alpha}--\ref{fig:fig_alpha_simpson}), though this difference was less drastic than in the diarrheal studies where all patients had active diarrhea.

@@

Finally, we sought to determine whether a unified microbiome response to general health and disease could be identified.
Previous studies have proposed that reduced alpha diversity is a reliable indicator of disease-associated dysbiosis \cite{cdi-vincent,ob-gordon,mosca2016gut}.
In our re-analysis, we found no consistent reduction of alpha diversity in case patients, with the exception of diarrhea and perhaps IBD (Supplementary Figures \ref{fig:fig_alpha}--\ref{fig:fig_alpha_simpson}).
These results are consistent with previous meta-analyses, which found inconsistent relationships between alpha diversity and disease and very small effect sizes in non-diarrheal diseases \cite{walters2014meta, Sze07092016}.
To further address the question of whether we could find a robust, generalized signal for diseased microbiomes regardless of the disease type, we built random forest classifiers to distinguish healthy patients from any type of case patient.
The AUCs from these general healthy vs. disease classifiers correlated strongly with the original single-dataset classification results, indicating that there is indeed a general microbiome signal that can be identified even across different diseases (see Supplementary Note \ref{sec:overall_classifier} and Supplementary Figure \ref{fig:overall_classifier}).

@@

Meta-analysis identified potential false positives and false negatives across studies and conditions.
For example, we found that reported associations between alpha diversity and disease within individual studies tended to lose significance when looking across studies, except in the case of diarrhea and perhaps IBD (Supplementary Figures \ref{fig:fig_alpha}--\ref{fig:fig_alpha_simpson}).
Another example of a potential false positive was the association between \textit{Prevotella} and disease.
Autism \cite{asd-kb}, rheumatoid arthritis \cite{ra-littman}, and HIV \cite{lozupone2013alterations,hiv-dinh} have each been reported to be associated with \textit{Prevotella}. For each of these diseases, the associations with \textit{Prevotella} were weakly significant or complicated by confounding factors.
In our more statistically conservative re-analysis, we found no association between autism or arthritis and \textit{Prevotella}.
As mentioned previously, in the case of HIV, the association with \textit{Prevotella} was due to demographic factors unrelated to disease \cite{noguera2016gut}.
Regardless of whether shifts in \textit{Prevotella} are truly biologically related to each studied disease state, it is clear that such shifts are not specific to one particular condition and should not be reported as putative disease-specific biomarkers.
We also found that certain signals picked out by meta-analysis did not always hold within individual studies.
For example, studies with small sample sizes often had few or no significant associations (e.g. Vincent et al. (2014) \cite{cdi-vincent}, Chen et al. (2012) \cite{crc-xiang}, and Willing et al. (2009) \cite{ibd-engstrand}).
Here, the fact that other studies analyzing the same diseases consistently found associations strengthens the hypothesis that the lack of microbiome-associated signal in these studies was due to low power rather than a lack of true signal.
Because individual studies are plagued by low statistical power, confounding variables, and batch effects which can obscure biological signals, the identification of disease-specific and non-specific microbial associations will continue to improve as more datasets and diseases are included in future meta-analyses.

@@

\subsection{16S processing}

Raw data were downloaded and processed through our in-house 16S processing pipeline ({\url{ https://github.com/thomasgurry/amplicon_sequencing_pipeline}).
Data and metadata were acquired as described in Supplementary Table \ref{tab:data}.
When needed, we de-multiplexed sequences by finding exact matches to the provided barcodes and trimmed primers with a maximum of 1 mismatch.
In general, sequences were quality filtered by truncating at the first base with Q $<$ 25.
However, some datasets did not pass this stringent quality threshold (i.e. the resulting OTU table was either missing many of the original samples, or the read depth was significantly lower than reported in the original paper).
For 454 data, we loosened the quality threshold to 20, whereas for paired-end Illumina data we removed reads with more than 2 expected errors.
If possible, all reads were trimmed to 200 bp.
In cases where this length trimming discarded a majority of sequences, we lowered our threshold to 150 or 101 bp.
The specific processing parameters we used for each dataset can be found in Supplementary Table \ref{tab:processing}.
To assign OTUs, we clustered OTUs at 100\% similarity using USEARCH \cite{edgar-usearch-2010} and assigned taxonomy to the resulting OTUs with the RDP classifier \cite{wang2007naive} and a confidence cutoff of 0.5.
For each dataset, we removed samples with fewer than 100 reads and OTUs with fewer than 10 reads, as well as OTUs which were present in fewer than 1\% of samples within a study.
We calculated the relative abundance of each OTU by dividing its value by the total reads per sample.
We then collapsed OTUs to genus level by summing their respective relative abundances, discarding any OTUs which were unannotated at the genus level.
All statistical analyses were performed on this genus-level relative abundance data.

@@

\FloatBarrier
\clearpage
\section{Figures}

\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=1\textwidth]{samplesize_auc_nsig_balance.pdf}}
    \captionsetup{font=footnotesize,labelfont=footnotesize}
	\caption{\textbf{Most diseases show microbiome alterations, and consistent disease-associated shifts differ in their extent and direction.} (A) Left: Total sample size for each study included in these analyses. Additional information about each dataset can be found in Table \ref{tab:datasets}. Studies on the y-axis are grouped by disease and ordered by decreasing sample size (top to bottom). Right: Area under the ROC curve (AUC) for genus-level random forest classifiers. X-axis starts at 0.5, the expected value for a classifier which assigns labels randomly, and AUCs less than 0.5 are not shown. ROC curves for all datasets are in Supplementary Figure \ref{fig:roc_curves}. Note that Youngster et al. (2014) \cite{cdi-youngster} had only 4 distinct control patients was excluded from the Random Forest analysis. (B) Left: Number of genera with q $<$ 0.05 (Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction) for each dataset. If a study has no significant associations, no point is shown. Right: Direction of the microbiome shift, i.e. the percent of total associated genera which were enriched in diseased patients. In datasets on the leftmost blue line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are health-associated (i.e. depleted in patients relative to controls). In datasets on the rightmost red line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are disease-associated (i.e. enriched in patients relative to controls). Supplementary Figures \ref{fig:overall_heatmap_qvalues} and \ref{fig:overall_heatmap_foldchange} show q-values and effects for each genus in each study.}
	\label{fig:fig1}
	\end{center}
\end{figure}
